\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2023-09-01 docs/papers/sto/so.tex
FIXUP draft notes

@@

% Main text: up to 3500 words: introduction + results + discussion
% so we should keep introduction and discussion brief
\vspace{1em}
\section{Introduction}
% without heading
Keywords: AI for science, differentiable simulations, deep neural networks,
symbolic interpretability

Abbrs: vector-Jacobian product (VJP), automatic differentiation (AD),

@@

\subsection{SO neural networks and symbolic expressions}

\subsection{Dark matter fields}

\subsection{Halo catalogs}

@@

Following these two works, we denote the particles' state with $\vz = (\vx,
\vp)^\intercal$ and their adjoint variables with $\vlambda = (\vxi, \vpi)$.
We refer to the quantities that are directly related to the final objective
function $\cJ$ as the observables.
In this work, our observable $\vo$ includes multiple snapshots at given times.
In the observation operator $O_{i}^{i-1}$, the adjoint variables are updated
through
\begin{align}
  \vxi_{i-1} &:= \vxi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vx_{i-1}}, \nonumber\\
  \vpi_{i-1} &:= \vpi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vp_{i-1}},
\end{align}
where the derivatives of $\vo$ with respect to $\vx_{i-1}$ and $\vp_{i-1}$ would
be non-zero if they were used in the interpolation of any snapshots in $\vo$.
The initial conditions are
\begin{align}
  \vxi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vx_n}, \nonumber\\
  \vpi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vp_n},
\end{align}
which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$.
The derivative of the objective function with respect to the observable and the
VJP of the observation funciton (mainly the interpolation function discussed
below) can be conveniently computed by AD.

>>>>

\item \YZ{Do we also consider pruning and distilling of the trained networks?}

@@

% Main text: up to 3500 words: introduction + results + discussion
% so we should keep introduction and discussion brief
\vspace{1em}
\section{Introduction}
% without heading
Keywords: AI for science, differentiable simulations, deep neural networks,
symbolic interpretability, fast cosmological simulations

Abbrs: vector-Jacobian product (VJP), automatic differentiation (AD), neural
network (NN)

% introduce PM and its limitation on small scales
Using a hybrid embedding of neural networks in differentiable physical
simulations, we improve the performance of PM method on small scales in
cosmological $N$-body simulations by sharpening the gravitational force.

@@

\subsection{SO Neural Networks and Symbolic Expressions}

% visualization & comparison of neural nets & symbolic functions
For the $f(k_i; \vvartheta)$ NN, we show it as a function of $k_i$ at multiple
different $\vvartheta$ values.

For the $g(k_1,k_2,k_3;\vvartheta)$ NN, which is permutationally symmetric about
$(k_1, k_2, k_3)$, we show it as a function of $k_i$ along axis and face
diagonal directions.

\subsection{Dark Matter Density Fields}

% transfer function & correlation coefficient for different confs

% density field with zoom in (with halos painted as circles?)

\subsection{Dark Matter Halo Catalogs}

@@

Following these two works, we denote the particles' state with $\vz = (\vx,
\vp)^\intercal$ and their adjoint variables with $\vlambda = (\vxi, \vpi)$.
We refer to the quantities that are directly related to the final objective
function $\cJ$ as the observables.
In this work, our observable $\vo$ includes multiple snapshots at given times.
For the observation operator $O_{i}^{i-1}$ in the adjoint equations from time
step $a_i$ back to $a_{i-1}$, the adjoint variables are updated through
\begin{align}
  \vxi_{i-1} &:= \vxi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vx_{i-1}}, \nonumber\\
  \vpi_{i-1} &:= \vpi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vp_{i-1}},
\end{align}
where the derivatives of $\vo$ with respect to $\vx_{i-1}$ and $\vp_{i-1}$ would
be non-zero if they were used in the interpolation of any snapshots in $\vo$.
The initial conditions are
\begin{align}
  \vxi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vx_n}, \nonumber\\
  \vpi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vp_n},
\end{align}
which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$.
The derivative of the objective function with respect to the observable and the
VJP of the observation funciton (mainly the interpolation function discussed
below) can be conveniently computed by AD.
